Inferring Galactic Parameters from Chemical Abundances with Simulation-Based Inference

### 2025-02-19 src/tex/ms.tex
[showyourwork] overleaf sync

   \author{Tobias Buck\inst{1,2}
          \and
          Berkay Günes\inst{1,2}
          \and
          Giuseppe Viterbo\inst{1,2}
          \and
          William H. Oliver\inst{1,2}
          \and
          et al.
          }

   \institute{Interdisciplinary Center for Scientific Computing (IWR), University of Heidelberg,
 Im Neuenheimer Feld 205, D-69120 Heidelberg, Germany
 \and
 Universität Heidelberg, Zentrum für Astronomie, Institut für Theoretische Astrophysik, Albert-Ueberle-Straße 2, D-69120 Heidelberg, Germany\\
 \email{tobias.buck@iwr.uni-heidelberg.de}
             }

@@

Recent advances in stellar spectroscopic surveys, such as APOGEE and GALAH, have produced unprecedented datasets of stellar chemical abundances. These datasets hold the potential to unlock detailed constraints on galactic parameters across diverse environments. However, traditional Bayesian inference methods, such as Markov Chain Monte Carlo (MCMC) and Hamiltonian Monte Carlo (HMC), struggle to scale to these large datasets. Such methods are computationally expensive, requiring hours of runtime for even modest sample sizes, and are susceptible to biases when confronted with high-dimensional posterior distributions.

@@

This study focuses on two critical global galactic parameters: 
the high-mass slope of the \citet[Tab.\,1]{2003PASP..115..763C} IMF ($\alpha_{\rm IMF}$) and the SN\,Ia normalization ($\log_{10}(N_{\rm Ia})$), the rate of SN\,Ia explosions per unit mass. We demonstrate the accuracy, scalability, and robustness of SBI through tests on mock datasets generated by \texttt{CHEMPY} \citep{Rybizki_2017}, as well as on data from hydrodynamical simulations. Additionally, we compare our results to those obtained using HMC-based inference on the same datasets \citep[see][]{Philcox_2019}, highlighting SBI's superior performance in terms of speed, precision, and resilience to model misspecification.

@@

@@

\begin{figure}[]
     \centering
     \includegraphics[width=\columnwidth]{figures/ape_NN.pdf}
     \vspace{-.5cm}
     \caption{Cumulative absolute percentage error of the NN emulator for the \texttt{CHEMPY} simulator. The orange histogram shows the cumulative distribution of percentage errors with the vertical dashed line indicating the median and the vertical dotted lines indicating the first and third quartile. The box plot on the top of the plot extends from the first quartile to the third quartile of the data, with a line at the median. The whiskers extend from the box to the farthest data point lying within $1.5\times$ the inter-quartile range from the box. The NN predicts abundances with an absolute percentage error far below typical observational errors.}
     \label{fig:ape_NN}
     \script{train_torch_chempy.py}
\end{figure}

@@

We use our SBI method described in the previous section to infer the global galactic parameters $\alpha_{\rm {IMF}}$ and $\log_{10}(N_{\rm{Ia}})$. In order to demonstrate the performance and robustness of our methods we use three mock data-sets: 
\begin{enumerate}
    \item Mock observations drawn from \texttt{CHEMPY} from the same yield set as the training data for the neural network emulator. With this we ensure to test our training strategy and the performance without any systematic distribution shifts. 
    \item  \texttt{CHEMPY} mock data using a different yield set to test for potential biases through model missspecification in our SBI setup.
    \item Simulated data from a large-scale hydrodynamical simulation taken from the IllustrisTNG suite \citep{Pillepich2018} but with the same yield set as our \texttt{CHEMPY} training data to ensure that we recover the correct parameters even for models with a completely different treatment of ISM physics.
\end{enumerate}

@@

For each individual observation consisting of the abundance measurements of a single star we sample the posterior estimate for all six parameters $\{\vec\Lambda,\vec\Theta_i,T_i\}$ with 1,000 points. This takes around $0.3s$ for each star, making it extremely fast to infer the parameters of a large amount of stars. Our method takes around $50$ seconds to build a posterior for all six parameters for a dataset size of 1,000 stars (each time sampling the single star posterior with 1,000 points and fitting for the Gaussian parameters). Our combined runtime for the NPE training plus sampling is then 4250 (2160 on Apple M! Max) seconds for 1,000 stars hence our methods making it more than $170$ ($340$) times faster than current HMC methods \citep[cf.][who need $40$h for only 200 stars]{Philcox_2019}. But since our approach is amortized and we do not need to retrain our NPE model each time we want to make an inference, we in fact achieve an inference speed-up of a factor of $\sim14,400$.
Hence, shorter computing times make it feasible to use orders of magnitudes more observations.

\paragraph{Validation through absolute percentage error}
We start evaluating our method by using \texttt{CHEMPY} to produce a mock observational dataset to ensure no systematical shft between training and testing data in terms of physics parameters.

To evaluate our method quantitatively, we compare the posterior mean to the ground truth value for each observation and calculate the absolute percentage error (APE, see Fig.~\ref{fig:posterior_APE}). For a single star observation, our NPE has an APE of \variable{output/global_posterior_APE.txt} when looking at all 6 parameters of interest. If we restrict ourselves to only the global two parameters, $\vec\Lambda = \{\alpha_\mathrm{IMF},\log_{10}(N_\mathrm{Ia})\}$, our NPE achieves an APE of \variable{output/posterior_APE.txt} as we show in Fig.~\ref{fig:posterior_APE}. Looking at Fig.~\ref{fig:posterior_APE} we find that for an individual star the accuracy of the NPE is not particularly high. However, for the global parameters $\vec\Lambda = \{\alpha_\mathrm{IMF},\log_{10}(N_\mathrm{Ia})\}$ we can boost the accuracy by combining the inference for many stars of the same galaxy.

@@

\textcolor{red}{here goes the assumption that stars are iid since each star is produced by a single call to \texttt{CHEMPY}. However, real physics might correlate stars abundances by star formation histories and cumulative enrichment...}

>>>>

   \author{Tobias Buck\inst{1,2}
          \and
          Berkay Günes\inst{1,2}
          \and
          Giuseppe Viterbo\inst{1,2}
          \and
          William H. Oliver\inst{1,2}
          \and
          Sven Buder\inst{3,4}
          }

   \institute{Interdisciplinary Center for Scientific Computing (IWR), University of Heidelberg,
 Im Neuenheimer Feld 205, D-69120 Heidelberg, Germany
 \and
 Universität Heidelberg, Zentrum für Astronomie, Institut für Theoretische Astrophysik, Albert-Ueberle-Straße 2, D-69120 Heidelberg, Germany
 \and
 Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia
 \and
 ARC Centre of Excellence for All Sky Astrophysics in 3 Dimensions (ASTRO 3D), Australia\\
 \email{tobias.buck@iwr.uni-heidelberg.de}
             }

@@

Recent advances in stellar spectroscopic surveys, such as APOGEE \citep{apogee17} and GALAH \citep{Buder2021,Buder2024}, have produced unprecedented datasets of stellar chemical abundances across a third of the period table. These datasets hold the potential to unlock detailed constraints on galactic parameters across diverse environments. However, traditional Bayesian inference methods, such as Markov Chain Monte Carlo (MCMC) and Hamiltonian Monte Carlo (HMC), struggle to scale to these large datasets. Such methods are computationally expensive, requiring hours of runtime for even modest sample sizes, and are susceptible to biases when confronted with high-dimensional posterior distributions.

@@

This study focuses on two critical global galactic parameters: 
the high-mass slope of the \citet[Tab.\,1]{2003PASP..115..763C} IMF ($\alpha_{\rm IMF}$) and the SN\,Ia normalization, $\log_{10}(N_{\rm Ia})$, the rate of SN\,Ia explosions per unit mass. We demonstrate the accuracy, scalability, and robustness of SBI through tests on mock datasets generated by \texttt{CHEMPY} \citep{Rybizki_2017}, as well as on data from hydrodynamical simulations. Additionally, we compare our results to those obtained using HMC-based inference on the same datasets \citep[see][]{Philcox_2019}, highlighting SBI's superior performance in terms of speed, precision, and resilience to model misspecification.

@@

\begin{table}[]
\caption{Alternative nucleosynthetic yield tables used in section \ref{sec: Results} for model misspecification tests.}
     \centering
     \begin{tabular}{c|c}
       Type & Yield Table \\
        \hline
         SN\,Ia & \citet{2003NuPhA.718..139T}\\
         %SN\,II & \citet{2013ARA&A..51..457N}\\
         AGB & \citet{2016ApJ...825...26K}
     \end{tabular}
\label{tab:chempy_ALT_yields}
\end{table}

@@

\begin{figure}[]
     \centering
     \includegraphics[width=\columnwidth]{figures/ape_NN.pdf}
     \vspace{-.5cm}
     \caption{Cumulative absolute percentage error of the NN emulator for the \texttt{CHEMPY} simulator. The orange histogram shows the cumulative distribution of percentage errors with the vertical dashed line indicating the median and the vertical dotted lines indicating the first and third quartile. The box plot on the top of the plot extends from the first quartile to the third quartile of the data, with a line at the median. The whiskers extend from the box to the farthest data point lying within $1.5\times$ the inter-quartile range from the box. The NN predicts abundances with an absolute percentage error far below typical observational errors.}
     \label{fig:ape_NN}
     \script{evaluate_emulator.py}
\end{figure}

@@

We use our SBI method described in the previous section to infer the global galactic parameters $\vec\Lambda = \{\alpha_\mathrm{IMF},\log_{10}(N_\mathrm{Ia})\}$. In order to demonstrate the performance and robustness of our methods we use three mock data-sets: 
\begin{enumerate}
    \item Mock observations drawn from \texttt{CHEMPY} from the same yield set as the training data for the neural network emulator. With this we ensure to test our training strategy and the performance without any systematic distribution shifts. 
    \item  \texttt{CHEMPY} mock data using a different yield set to test for potential biases through model misspecification in our SBI setup. (see  \ref{tab:chempy_ALT_yields})
    \item Simulated data from a large-scale hydrodynamical simulation taken from the IllustrisTNG suite \citep{Pillepich2018} but with the same yield set as our \texttt{CHEMPY} training data to ensure that we recover the correct parameters even for models with a completely different treatment of ISM physics.
\end{enumerate}

@@

For each individual observation consisting of the abundance measurements of a single star we sample the posterior estimate for all six parameters $\{\vec\Lambda,\vec\Theta_i,T_i\}$ with 1,000 points. This takes around $0.3s$ for each star, making it extremely fast to infer the parameters of a large amount of stars. Our method takes around $50$ seconds to build a posterior for all six parameters for a dataset size of 1,000 stars (each time sampling the single star posterior with 1,000 points and fitting for the Gaussian parameters). Our combined runtime for the NPE training plus sampling is then $2160$ seconds on an Apple M1 Max for 1,000 stars hence our methods making it more than $340$ times faster than current HMC methods which take around $40$h for only $200$ stars \citep{Philcox_2019}. But since our approach is amortized and we do not need to retrain our NPE model each time we want to make an inference, we in fact achieve an inference speed-up of a factor of $\sim14,400$.
Hence, shorter computing times make it feasible to use orders of magnitudes more observations.

\paragraph{Validation through absolute percentage error}
We start evaluating our method by using \texttt{CHEMPY} to produce a mock observational dataset to ensure no systematical shift between training and testing data in terms of physics parameters.

To evaluate our method quantitatively, we compare the posterior mean to the ground truth value for each observation and calculate the absolute percentage error (APE, see Fig.~\ref{fig:posterior_APE}). For a single star observation, our NPE has an APE of \variable{output/global_posterior_APE.txt} when looking at all 6 parameters of interest.
If we restrict ourselves to only the two global parameters $\vec\Lambda$, our NPE achieves an APE of \variable{output/posterior_APE.txt} as shown in Fig.~\ref{fig:posterior_APE}. There we can also see, that the accuracy for an individual star of the NPE is not particularly high. However, for the global parameters $\vec\Lambda$ we can boost the accuracy by combining the inference for many stars of the same galaxy.

@@

%\textcolor{red}{here goes the assumption that stars are iid since each star is produced by a single call to \texttt{CHEMPY}. However, real physics might correlate stars abundances by star formation histories and cumulative enrichment...}
